New Metrics for Assessing Projection Pursuit Indexes, and Guiding Optimisation Choices

### 2024-09-11 jso.qmd
small changes to table 1

<!-- **Projection distance**: The projection distance between projection -->

<!-- bases and the optimal best basis (or the empirical best basis if -->

<!-- the optimal is unknown) measures how close a visited basis is to  -->

<!-- the best basis and it is calculated as the Frobenius norm in equation -->

<!-- (@eq-proj-dist). Plotting projection distances against index values for the  -->

<!-- bases visited reveals how the index value changes as the basis -->

<!-- approaches the optimal basis, helping to calculate the squintability -->

<!-- measure. -->

@@

The index properties, including smoothness and squintability, offer
numerical metrics to characterise the complexity of projection pursuit
optimisation problems. @tbl-smoothness-squintability presents the
parameters for calculating both metrics estimated from the Gaussian
process (variance, range, smooth, and nugget) and the scaled logistic
function ($\theta_1$ to $\theta_4$) for each case considered in the
simulation. The column "smooth" is used as the smoothness metric and the
column "squint" is calculated as equation (@eq-squintability-parametric)
as the squintability metric. @tbl-mod-output presents the results of
fitting a generalised linear model with a binomial family and a logit
link function.<!--, where all the three
components (success rate, JS hyper-parameters, and index properties) are
combined.--> The model suggests that JSO success rate is positively
associated with the two hyper-parameters, as well as with the index
properties: smoothness and squintability. Specifically, using 10 more
jellyfish and 10 more tries increases the odd ratio of success by
`r round(100 * (exp(broom::tidy(mod1)$estimate[6])-1), 2)`% and
`r round(100 * (exp(broom::tidy(mod1)$estimate[7])-1), 2)`%,
respectively. However, being flagged with long runtime and an increase
of data dimension reduce the success rate by
`r round(exp(broom::tidy(mod1)$estimate[5]) * 100, 2)`% and
`r round(exp(broom::tidy(mod1)$estimate[4]) * 100, 2)`%, respectively.
The variable `squintability` and `dimension` are significant, suggesting
their importance relative to JSO hyper-parameters in the optimisation
success.

@@

```{r}
#| label: tbl-smoothness-squintability
#| tbl-cap: Parameters estimated from the Gaussian process (outputscale $\eta$, lengthscale $\ell$, smoothness $\nu$, and nugget $\sigma$) and scaled logistic function ($\theta_1$ to $\theta_4$ and $\varsigma$) for the pipe-finding and sine-wave finding problems.  The columns $\nu$ and $\varsigma$ represent the smoothness and squintability measures respectively. 
dt <- tibble(shape = c(rep("pipe", 4), rep("sine", 8)), smoothness) |>
  left_join(squintability) |> 
  rename(d = n, smooth = smoothness) |> 
  mutate(index = ifelse(index == "dcor2d_2", "dcor2d", index)) 
dt |> 
  knitr::kable(
    digits = 2, format = "latex", 
    col.names = c(
      "shape", "index", "d",
      "$\\eta$", "$\\ell$", "$\\nu$", "$\\sigma$",
      "$\\theta_1$", "$\\theta_2$", "$\\theta_3$", "$\\theta_4$", "$\\varsigma$"), linesep = "",
    booktabs = T, escape = FALSE, row.names = TRUE) |> 
  column_spec(1, border_left = TRUE) |> 
  column_spec(c(4, 8, 13), border_right = TRUE) |> 
  kable_styling(font_size = 10) 
```

<!-- 
```{r}
#| label: tbl-mod-data
#| tbl-cap: Data constructed to store simulation results (the first 7 rows) includes name of index, dimension, index metrics, JSO parameters, and performance metrics. 
sim_df |> 
  dplyr::select(index, d, smoothness, 
    squintability, n_jellies, max_tries, 
    -I_max, P_J, time) |> 
  mutate(time = as.numeric(time)) |>
  arrange(index, d, n_jellies, max_tries) |> 
  head(7) |> 
  kable(digits = 2, format = "latex", 
        col.names = c("index", "d", 
          "smoothness", "squintability", 
          "n. jellyfish", 
          "max. tries", 
          "success rate", "time (sec)"),
        linesep = "",
        booktabs = T) |> 
  kable_styling(#latex_options = "hold_position",
    font_size = 10)
```
-->

```{r}
#| label: tbl-mod-output
#| tbl-cap: Model estimates of proportion of jellyfish success on index properties and jellyfish hyper-parameters from the generalised linear model with a binomial family and a logit link function. The variable smoothness, squintability, number of jellyfish and maximum number of tries are positively associated with JSO success rate while data dimension and being flagged as long runtime are negatively associated with the success rate. The variable squintability and dimension are significant, suggesting their importance relative to jellyfish hyper-parameters in the optimisation success.
broom::tidy(mod1) |> 
  mutate(term = c("Intercept", "Smoothness",
   "Squintability", "Dimension (d)", 
   "Long time", "N. jellyfish", 
   "Max. tries")) |>
  kable(digits = 2, format = "latex", 
    linesep = "", booktabs = T) |> 
  column_spec(c(1,5), border_right = TRUE) |> 
  column_spec(1, border_left = TRUE) |> 
  kable_styling(font_size = 10) 
```

>>>>

@@

The index properties, including smoothness and squintability, offer
numerical metrics to characterise the complexity of projection pursuit
optimisation problems. @tbl-smoothness-squintability presents the
parameters for calculating both metrics estimated from the Gaussian
process (variance, range, smooth, and nugget) and the scaled logistic
function ($\theta_1$ to $\theta_4$) for each case considered in the
simulation. The column "smooth" is used as the smoothness metric and the
column "squint" is calculated as equation (@eq-squintability-parametric)
as the squintability metric. @tbl-mod-output presents the results of
fitting a generalised linear model with a binomial family and a logit
link function. The model suggests that JSO success rate is positively
associated with the two hyper-parameters, as well as with the index
properties: smoothness and squintability. Specifically, using 10 more
jellyfish and 10 more tries increases the odd ratio of success by
`r round(100 * (exp(broom::tidy(mod1)$estimate[6])-1), 2)`% and
`r round(100 * (exp(broom::tidy(mod1)$estimate[7])-1), 2)`%,
respectively. However, being flagged with long runtime and an increase
of data dimension reduce the success rate by
`r round(exp(broom::tidy(mod1)$estimate[5]) * 100, 2)`% and
`r round(exp(broom::tidy(mod1)$estimate[4]) * 100, 2)`%, respectively.
The variable `squintability` and `dimension` are significant, suggesting
their importance relative to JSO hyper-parameters in the optimisation
success.

@@

```{r}
#| label: tbl-smoothness-squintability
#| tbl-cap: Parameters estimated from the Gaussian process (outputscale $\eta$, lengthscale $\ell$, smoothness $\nu$, and nugget $\sigma$) and scaled logistic function ($\theta_1$ to $\theta_4$ and $\varsigma$) for the pipe-finding and sine-wave finding problems.  The columns $\nu$ and $\varsigma$ represent the smoothness and squintability measures respectively. 
dt <- tibble(shape = c(rep("pipe", 4), rep("sine", 8)), smoothness) |>
  left_join(squintability) |> 
  rename(d = n, smooth = smoothness) |> 
  mutate(index = ifelse(index == "dcor2d_2", "dcor", index),
         index = ifelse(index == "loess2d", "loess", index),
         index = ifelse(index == "splines2d", "splines", index),
         variance = sqrt(variance)) 
dt |> 
  knitr::kable(
    digits = 2, format = "latex", 
    col.names = c(
      "shape", "index", "d",
      "$\\eta$", "$\\ell$", "$\\nu$", "$\\sigma$",
      "$\\theta_1$", "$\\theta_2$", "$\\theta_3$", "$\\theta_4$", "$\\varsigma$"), linesep = "",
    booktabs = TRUE, escape = FALSE, row.names = TRUE, align = "c") |> 
  column_spec(1, border_left = TRUE) |> 
  column_spec(c(4, 8, 13), border_right = TRUE) |> 
  kable_styling(font_size = 10) 
```

```{r}
#| label: tbl-mod-output
#| results: "asis"
#| tbl-cap: Model estimates of proportion of jellyfish success on index properties and jellyfish hyper-parameters from the generalised linear model with a binomial family and a logit link function. The variable smoothness, squintability, number of jellyfish and maximum number of tries are positively associated with JSO success rate while data dimension and being flagged as long runtime are negatively associated with the success rate. The variable squintability and dimension are significant, suggesting their importance relative to jellyfish hyper-parameters in the optimisation success.
broom::tidy(mod1) |> 
  mutate(term = c("Intercept", "Smoothness",
   "Squintability", "Dimension (d)", 
   "Long time", "N. jellyfish", 
   "Max. tries")) |>
  kable(digits = 2, format = "latex", 
    linesep = "", booktabs = T) |> 
  column_spec(c(1,5), border_right = TRUE) |> 
  column_spec(1, border_left = TRUE) |> 
  kable_styling(font_size = 10) 
```

### 2024-09-11 jso.tex
small changes to table 1

\begin{tcolorbox}[enhanced jigsaw, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colback=white, opacitybacktitle=0.6, left=2mm, titlerule=0mm, toprule=.15mm, arc=.35mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, leftrule=.75mm, colframe=quarto-callout-note-color-frame, opacityback=0, breakable, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm]

@@

\begin{tabular}{|>{}lll>{}r|rrr>{}r|rrrr>{}r|}
\toprule
  & shape & index & d & $\eta$ & $\ell$ & $\nu$ & $\sigma$ & $\theta_1$ & $\theta_2$ & $\theta_3$ & $\theta_4$ & $\varsigma$\\
\midrule
1 & pipe & holes & 6 & 0.00 & 0.37 & 2.36 & 0.22 & 1.02 & 0.86 & 3.37 & 0.02 & 3.05\\
2 & pipe & holes & 8 & 0.00 & 0.18 & 2.19 & 0.82 & 1.01 & 0.87 & 3.26 & 0.03 & 2.96\\
3 & pipe & holes & 10 & 0.00 & 0.11 & 2.19 & 3.03 & 1.02 & 0.88 & 3.15 & 0.02 & 2.95\\
4 & pipe & holes & 12 & 0.00 & 0.15 & 2.29 & 1.58 & 1.01 & 0.88 & 3.34 & 0.00 & 3.12\\
5 & sine & MIC & 6 & 0.02 & 0.09 & 2.46 & 0.08 & 0.89 & 0.57 & 1.62 & -0.02 & 1.26\\
6 & sine & MIC & 8 & 0.02 & 0.08 & 2.64 & 0.08 & 0.93 & 0.33 & 1.31 & -0.03 & 0.77\\
7 & sine & TIC & 6 & 0.12 & 0.11 & 2.44 & 0.09 & 0.95 & 0.54 & 1.72 & -0.03 & 1.32\\
8 & sine & TIC & 8 & 0.12 & 0.10 & 2.47 & 0.09 & 0.95 & 0.56 & 1.72 & -0.03 & 1.37\\
9 & sine & dcor2d & 6 & 0.03 & 0.17 & 2.66 & 0.11 & 0.95 & 1.04 & 2.74 & -0.02 & 2.95\\
10 & sine & loess2d & 6 & 0.08 & 0.34 & 1.99 & 0.31 & 1.02 & 1.04 & 2.65 & 0.08 & 2.76\\
11 & sine & splines2d & 6 & 0.04 & 0.24 & 2.54 & 0.10 & 1.01 & 1.05 & 2.73 & -0.01 & 3.12\\
12 & sine & stringy & 6 & 0.00 & 0.01 & 1.54 & 38.17 & 1.05 & 0.01 & 254.73 & 0.05 & 2.96\\
\bottomrule
\end{tabular}
\endgroup{}

@@

The article is created using Quarto (Allaire et al. 2022) in R (R Core
Team 2023). The source code for reproducing the work reported in this
paper can be found at:
\url{https://github.com/huizezhang-sherry/paper-jso}. The simulation
data produced as described in Section 5 in can be found at
\url{https://figshare.com/articles/dataset/Simulated_raw_data/26039506}.
Nicolas Langrené acknowledges the partial support of the Guangdong
Provincial Key Laboratory IRADS (2022B1212010006, R0400001-22) and the
UIC Start-up Research Fund UICR0700041-22.

>>>>

\begin{tcolorbox}[enhanced jigsaw, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, leftrule=.75mm, rightrule=.15mm, bottomrule=.15mm, bottomtitle=1mm, toptitle=1mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, opacitybacktitle=0.6, breakable, arc=.35mm, titlerule=0mm, toprule=.15mm, opacityback=0, left=2mm, colframe=quarto-callout-note-color-frame, colback=white]

@@

\begin{tabular}{|>{}lcc>{}c|ccc>{}c|cccc>{}c|}
\toprule
  & shape & index & d & $\eta$ & $\ell$ & $\nu$ & $\sigma$ & $\theta_1$ & $\theta_2$ & $\theta_3$ & $\theta_4$ & $\varsigma$\\
\midrule
1 & pipe & holes & 6 & 0.04 & 0.37 & 2.36 & 0.22 & 1.02 & 0.86 & 3.37 & 0.02 & 3.05\\
2 & pipe & holes & 8 & 0.01 & 0.18 & 2.19 & 0.82 & 1.01 & 0.87 & 3.26 & 0.03 & 2.96\\
3 & pipe & holes & 10 & 0.01 & 0.11 & 2.19 & 3.03 & 1.02 & 0.88 & 3.15 & 0.02 & 2.95\\
4 & pipe & holes & 12 & 0.01 & 0.15 & 2.29 & 1.58 & 1.01 & 0.88 & 3.34 & 0.00 & 3.12\\
5 & sine & MIC & 6 & 0.13 & 0.09 & 2.46 & 0.08 & 0.89 & 0.57 & 1.62 & -0.02 & 1.26\\
6 & sine & MIC & 8 & 0.13 & 0.08 & 2.64 & 0.08 & 0.93 & 0.33 & 1.31 & -0.03 & 0.77\\
7 & sine & TIC & 6 & 0.35 & 0.11 & 2.44 & 0.09 & 0.95 & 0.54 & 1.72 & -0.03 & 1.32\\
8 & sine & TIC & 8 & 0.35 & 0.10 & 2.47 & 0.09 & 0.95 & 0.56 & 1.72 & -0.03 & 1.37\\
9 & sine & dcor & 6 & 0.18 & 0.17 & 2.66 & 0.11 & 0.95 & 1.04 & 2.74 & -0.02 & 2.95\\
10 & sine & loess & 6 & 0.28 & 0.34 & 1.99 & 0.31 & 1.02 & 1.04 & 2.65 & 0.08 & 2.76\\
11 & sine & splines & 6 & 0.20 & 0.24 & 2.54 & 0.10 & 1.01 & 1.05 & 2.73 & -0.01 & 3.12\\
12 & sine & stringy & 6 & 0.00 & 0.01 & 1.54 & 38.17 & 1.05 & 0.01 & 254.73 & 0.05 & 2.96\\
\bottomrule
\end{tabular}
\endgroup{}

@@

The article is created using Quarto (Allaire et al. 2022) in R (R Core
Team 2023). The source code for reproducing the work reported in this
paper can be found at:
\url{https://github.com/huizezhang-sherry/paper-jso}. The simulation
data produced in Section 5 can be found at
\url{https://figshare.com/articles/dataset/Simulated_raw_data/26039506}.
Nicolas Langrené acknowledges the partial support of the Guangdong
Provincial Key Laboratory IRADS (2022B1212010006, R0400001-22) and the
UIC Start-up Research Fund UICR0700041-22.
